Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-11-15 manuscript/manuscript.Rmd
Cleaning up discussion, adding refs

Here, we saw how changes in the density surface driven by physics during this process led to the coupling of physics, chemistry, and biology. The enrichment of eukaryotic organisms at the DCM of the cyclone mirrors other ocean processes where upwelling of deep, nutrient-rich water leads to increases in nitrogen-rich metabolites representative of a shift towards organisms such as diatoms and dinoflagellates that undergo bloom dynamics [@Durham2022; @Kulisch2023]. @Barone2022 established a "chemical wake" hypothesis in which cyclone formation introduces a transient phytoplankton community with net photosynthetic production that induces export, a conceptual model that can be extended to include many of the metabolites named here, as some metabolites are more bioavailable than others and the "flavor" of this chemical wake will change over time.

Almost all large phytoplankton classes in the 2017 eddy pair were enriched in the cyclone DCM relative to the anticyclone (particularly prymnesiophytes, dinoflagellates, diatoms, and green algae), while *Prochlorococcus* was significantly less abundant and *Synechococcus* was approximately equal [@Barone2022]. Thus, the metabolite data presented here does not help distinguish between large phytoplankton classes but instead provides additional cues about the shift from a metabolome dominated by cyanobacteria producers into one dominated by eukaryotic organisms. This shift occurs anywhere

@@

Both cruises occurred in weakening cyclones and stable anticyclones

>>>>

Here, we saw how changes in the density surface driven by physics during this process led to the coupling of physics, chemistry, and biology. The enrichment of eukaryotic organisms at the DCM of the cyclone mirrors other ocean processes where upwelling of deep, nutrient-rich water leads to increases in nitrogen-rich metabolites representative of a shift towards organisms such as diatoms and dinoflagellates that undergo bloom dynamics [@Durham2022; @Kuhlisch2023]. @Barone2022 established a "chemical wake" hypothesis in which cyclone formation introduces a transient phytoplankton community with net photosynthetic production that induces export, a conceptual model that can be extended to include many of the metabolites named here, as some metabolites are more bioavailable than others and the "flavor" of this chemical wake will change over time.

@@

Almost all large phytoplankton classes in the 2017 eddy pair were enriched in the cyclone DCM relative to the anticyclone (particularly prymnesiophytes, dinoflagellates, diatoms, and green algae), while *Prochlorococcus* was significantly less abundant and *Synechococcus* was approximately equal [@Barone2022]. Thus, the metabolite data presented here does not help distinguish between large phytoplankton classes but instead provides additional cues about the shift from a metabolome dominated by cyanobacteria producers into one dominated by eukaryotic organisms. This shift occurs naturally over the life cycle of an eddy, with distinctions typically drawn between the eddy intensification, its mature phase, and eventual decay [@Rii2008; @McGillicuddy2016; @Kuhlisch2023; @Barone2022]. Both cruises occurred in weakening cyclones and mature anticyclones, limiting our ability to make direct conclusions about all eddy life cycle phases, but the general alignment of the results presented here with our initial hypotheses strengthens our confidence in metabolites as links between eddy biogeochemistry and their biological shifts anywhere.
